---
title: "Leafcutter Analysis: Hardy Bulk"
author: 
- name: "Jonathan Brenton & Guillermo Rocamora Pérez"
  affiliation: UCL - Version 1.4
output:
  bookdown::html_document2:
    figure_caption: yes
    code_folding: hide
    theme: paper
    highlight: haddock
    df_print: paged
    toc: true
    toc depth: 3
    toc_float: true
    number_sections: true
  md_document:
    variant: markdown_github
    toc: true
    number_sections: true
always_allow_html: true
link-citations: true
notes-after-punctuation: false
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/dependent/jhep-reports.csl
---

This is an adapated version of Guillermo's Hardy Limma analysis script.

*Notes*
Pathways with only 1 gene count in them were removed.

Similarity threshold of 0.7 was used to reduce the GO pathways.

# Setup
```{r setup, echo=F, eval=T, warning=F, message=FALSE}
library(clusterProfiler)
library(org.Hs.eg.db)
library(magrittr)
library(patchwork)
library(tidyverse)

knitr::opts_chunk$set(
 echo=F, eval=T, warning=F, message=FALSE
)




main_path<-"~/jbrenton/ASAP_Hardy_post_align_files/Final_Dataset/leafcutter/Final_Analysis/"
results_path<-("~/jbrenton/ASAP_Hardy_post_align_files/Final_Dataset/leafcutter/Final_Analysis/Results/")
full_results<-readRDS("/home/jbrenton/ASAP_Hardy_post_align_files/Final_Dataset/leafcutter/Final_Analysis/Data/PD_groups_collapsed_blist_removed/full_results_list_for_default_params.RDS")

sig_res_df<-full_results$all_results$significant_clusters_0.05_filter

sig_res_df$Test<-sig_res_df$comparison

sig_res_df %<>% mutate(brain_area=gsub(x = comparison, replacement = "\\1", pattern="^(.+)\\.[A-Z].*$")) %>%
  mutate(comparison=gsub(x = comparison, replacement = "\\1", pattern="^.+\\.([A-Z].*)$")) %>% dplyr::select(brain_area, comparison, everything())

all_clusters<-
  full_results$all_results$cluster_significance %>% 
    # filter for successful tests and remove clusters overlapping multiple genes
    dplyr::filter(status == "Success", !str_detect(genes, ",")) %>%
    tidyr::separate(cluster, into = c("chr", "cluster"), sep = ":") %>% 
    dplyr::inner_join(full_results$all_results$intron_usage %>% 
                        tidyr::separate(intron, into = c("chr", "start", "end", "cluster"), sep = ":")) %>% 
    dplyr::mutate(direction_of_effect = case_when(deltapsi > 0 ~ "upregulated",
                                                  deltapsi < 0 ~ "downregulated",
                                                  deltapsi == 0 ~ "no_change"))

universe <- all_clusters %>% 
  distinct(genes) %>% pull()

# Report parameters
count_threshold <- 2
go_reduce_threshold <- 0.7

```

```{r setup-theme, echo = F, results = "asis"}
## Custom sciRmdTheme with increased width.
## Please ask guillermorocamora@gmail.com for more details.
if(require(grpSciRmdTheme)){
  grpSciRmdTheme::set.theme(
    theme = "default",
    color = NULL,
    header.sticky = FALSE,
    list.group.icon = "arrow",
    font.family = "Arial",
    font.color = "black",
    header.color = "darkblue",
    figure.perc = 100
  )
}
## Custom styles for the output html

# Setting up tab bars
cat('
<style type="text/css">
.dataTables_scrollHeadInner{
  width:100% !important;
}
.dataTables_scrollHeadInner table{
  width:100% !important;
}
/*
.code-folding-btn {
  display: none;
}
*/
h3, .h3 {
  font-size: 22px!important;
}
h4, .h4 {
  font-size: 18px!important;
}
h5, .h5 {
  font-size: 16px!important;
}
body{
  font-size: 13px;
}

.tocify-subheader {
    text-indent: 15px;
    display: none;
    font-size: 12px;
}


.break5 .nav-tabs, .break4 .nav-tabs, .break3 .nav-tabs, .break2 .nav-tabs{
  display: grid;
  text-align: center;
  font-size: 14px;
  font-weight: bold;
}
.break5 .nav-tabs:before, .break4 .nav-tabs:before, .break3 .nav-tabs:before, .break2 .nav-tabs:before,
.break5 .nav-tabs:after, .break4 .nav-tabs:after, .break3 .nav-tabs:after, .break2 .nav-tabs:after {
  content: unset;
}

.break5 .nav-tabs{
  grid-template-columns: repeat(5, 1fr);
}
.break4 .nav-tabs{
  grid-template-columns: repeat(4, 1fr);
}
.break3 .nav-tabs{
  grid-template-columns: repeat(3, 1fr);
}
.break2 .nav-tabs{
  grid-template-columns: repeat(2, 1fr);
}

.breakB5 .nav-tabs{
  grid-template-columns: repeat(5, 1fr);
  font-size: 12px;
}
.breakB4 .nav-tabs{
  grid-template-columns: repeat(4, 1fr);
  font-size: 12px;
}
.breakB3 .nav-tabs{
  grid-template-columns: repeat(3, 1fr);
  font-size: 12px;
}
.breakB2 .nav-tabs{
  grid-template-columns: repeat(2, 1fr);
  font-size: 12px;
}

.stop-break .nav-tabs {
  display: flex;
  font-size: 13px;
  font-weight: normal;
}
.stop-break .nav-tabs:before,
.stop-break .nav-tabs:after {
  content: unset;
}
</style>')
```


## Load gene info

```{r load gene info, echo=F, eval=T, warning=F, message=FALSE}

# Load gene info
uniKeys <- AnnotationDbi::keys(org.Hs.eg.db, keytype = "ENSEMBL")
cols <- c("SYMBOL", "ENTREZID", "ENSEMBL")
cross_comp_list <- AnnotationDbi::select(org.Hs.eg.db, keys = uniKeys, columns = cols, keytype = "ENSEMBL")

# Remove duplicate symbols
non_duplicates_idx <- which(duplicated(cross_comp_list$SYMBOL) == F)
cross_comp_list <- cross_comp_list[non_duplicates_idx, ]
```


```{r generate cluster dataframes, echo=F, eval=T, warning=F, message=FALSE}

 universe <- AnnotationDbi::select(org.Hs.eg.db, keys=universe,
                                    columns=c("ENTREZID"), keytype="SYMBOL")

 universe<-na.omit(universe$ENTREZID)
  
  universe<-as.vector(universe)
  
  universe<-universe[which(duplicated(universe) == FALSE)]
  
  
  x<-sig_res_df
  x<-dplyr::rename(x, 'SYMBOL'=genes)
  
  y<-left_join(x, cross_comp_list, by="SYMBOL")
  
  z<-y %>% 
    group_by(brain_area) %>% 
    distinct(ENTREZID,.keep_all = T) %>% 
    ungroup()
  
  y<-z[-which(is.na(z$ENTREZID)),]
  
  count_threshold <- 2
```


## Run cluster profiler

```{r running clusterProfiler, echo=F, eval=T, warning=F, message=FALSE}
# running clusterProfiler

formula_res_merge_path <- file.path(results_path, "Final_Hardy_Leaf_formula_GO_merge.rds")
formula_res_merge_red_path <- file.path(results_path, "Final_Hardy_Leaf_formula_GO_merge_red.rds")

if(!file_test("-f", formula_res_merge_path, formula_res_merge_red_path)){

formula_res_mf <- clusterProfiler::compareCluster(
  ENTREZID ~ brain_area,
  data = y, 
  fun = enrichGO,
  OrgDb = org.Hs.eg.db, ont = "MF",
  universe = universe,
  readable = T) %>%
  dplyr::mutate(go_type = "MF", go_id = ID)

formula_res_bp <- clusterProfiler::compareCluster(
  ENTREZID ~ brain_area,
  data = y, 
  fun = enrichGO,
  OrgDb = org.Hs.eg.db, ont = "BP",
  universe = universe,
    readable = T) # is empty so not trying to run mutate on empty tibble as gives error
# %>%
  # dplyr::mutate(go_type = "BP", go_id = ID)

formula_res_cc <- clusterProfiler::compareCluster(
  ENTREZID ~ brain_area,
  data = y, 
  fun = enrichGO,
  OrgDb = org.Hs.eg.db, ont = "CC",
  universe = universe,
    readable = T) %>%
  dplyr::mutate(go_type = "CC", go_id = ID)

formula_res_mf_red <- formula_res_mf
formula_res_bp_red <- formula_res_bp
formula_res_cc_red <- formula_res_cc



# New function to go with updated rrvgo - previous problem mainly with reduce sim function - both score parameter and parent_sim_score are changed
# scores is now a character vector choice of either "size" or "uniqueness" - uniqueness is default
# parent sim score is now divided into term uniqueness and term dissimilaity or something
updated_go_reduce<-function (pathway_df, orgdb = "org.Hs.eg.db", threshold = 0.7, 
                             scores = NULL, measure = "Wang") 
{
  if (!measure %in% c("Resnik", "Lin", "Rel", "Jiang", "Wang")) {
    stop("Chosen measure is not one of the recognised measures, c(\"Resnik\", \"Lin\", \"Rel\", \"Jiang\", \"Wang\").")
  }
  if (measure == "Wang") {
    computeIC <- FALSE
  }
  else {
    computeIC <- TRUE
  }
  ont <- pathway_df %>% .[["go_type"]] %>% unique()
  if (any(!ont %in% c("BP", "CC", "MF"))) {
    stop("Column go_type does not contain the recognised sub-ontologies, c(\"BP\", \"CC\", \"MF\")")
  }
  go_similarity <- setNames(object = vector(mode = "list", 
                                            length = length(ont)), nm = ont)
  for (i in 1:length(ont)) {
    print(stringr::str_c("Reducing sub-ontology: ", ont[i]))
    hsGO <- GOSemSim::godata(OrgDb = orgdb, ont = ont[i], 
                             computeIC = computeIC)
    terms <- pathway_df %>% dplyr::filter(.data$go_type == 
                                            ont[i]) %>% .[["go_id"]] %>% unique()
    sim <- GOSemSim::mgoSim(GO1 = terms, GO2 = terms, semData = hsGO, 
                            measure = measure, combine = NULL)
    go_similarity[[i]] <- rrvgo::reduceSimMatrix(simMatrix = sim, 
                                                 threshold = threshold, orgdb = orgdb) %>% 
      tibble::as_tibble() %>% dplyr::rename(parent_id = .data$parent, 
                                            parent_term = .data$parentTerm, parent_sim_score = .data$termUniqueness)
  }
  go_sim_df <- go_similarity %>% qdapTools::list_df2df(col1 = "go_type")
  pathway_go_sim_df <- pathway_df %>% dplyr::inner_join(go_sim_df %>% 
                                                          dplyr::select(.data$go_type, go_id = .data$go, contains("parent")), 
                                                        by = c("go_type", "go_id")) %>% dplyr::arrange(.data$go_type, 
                                                                                                       .data$parent_id, -.data$parent_sim_score)
  return(pathway_go_sim_df)
}

# unable to reduce with one term but wanted to include it was a 1 ID term so added code to make the relevant columns
formula_res_mf_red@compareClusterResult %<>% mutate(parent_id=go_id, 
                                                   parent_term=Description,
                                                   parent_sim_score=1 # used 1 as arbitrary number
                                                   )
# bp empty so can't run the below:
# formula_res_bp_red@compareClusterResult %<>% updated_go_reduce(threshold = go_reduce_threshold)
formula_res_cc_red@compareClusterResult %<>% updated_go_reduce(threshold = go_reduce_threshold)

# Merge ontologies
formula_res_merge <- clusterProfiler::merge_result(list(mf = formula_res_mf,
                                                        bp = formula_res_bp,
                                                        cc = formula_res_cc))


formula_res_merge_red <- clusterProfiler::merge_result(list( # removed mf and bp since they can't be reduced
  mf = formula_res_mf_red,
  # bp = formula_res_bp_red,
  cc = formula_res_cc_red))

# Filter pathways
names(formula_res_merge@compareClusterResult)[1] <- "Ontology"
formula_res_merge@compareClusterResult %<>% dplyr::filter(Count >= count_threshold)

names(formula_res_merge_red@compareClusterResult)[1] <- "Ontology"
formula_res_merge_red@compareClusterResult %<>% dplyr::filter(Count >= count_threshold)


  # Save results
  formula_res_merge %>% saveRDS(formula_res_merge_path)
  formula_res_merge_red %>% saveRDS(formula_res_merge_red_path)
}else{
  formula_res_merge <- readRDS(formula_res_merge_path)
  formula_res_merge_red <- readRDS(formula_res_merge_red_path)
}



```

## Number of Pathways per Brain Area Table

Number of Clusters found per tissue and regulation shown below.

```{r Number of Cluster Table, echo=F, eval=T, warning=F, message=FALSE}

formula_res_merge@compareClusterResult %>% dplyr::count(brain_area) %>%
  tibble::as_tibble()
# %>%
#   tidyr::separate(col = "Cluster", into = c("Brain_area", "Regulation"), sep = "\\.", extra = "merge")

```

### Pathway List

```{r}

formula_res_merge@compareClusterResult %>% dplyr::select(brain_area, Description) %>% arrange(brain_area)
```


## Functions for Hardy data

The term reduction was performed across all brain areas/comparisons at the same time. Threshold was set at 0.7.

```{r}
generatePlotDataframe <- function(formula_res){
  if("parent_term" %in% colnames(formula_res@compareClusterResult)){
    description_names <- formula_res@compareClusterResult %>% 
      dplyr::distinct(Description, parent_term) %>% 
      dplyr::group_by(parent_term) %>% 
      dplyr::count() %>%
      dplyr::mutate(mod_description = stringr::str_to_title(paste0(parent_term, " (", n, ")"))) %>%
      dplyr::mutate(mod_description = paste0(strwrap(mod_description, width = 50, prefix = "\n", initial = ""),
                                             collapse = "")) %>%
      dplyr::select(parent_term, n, mod_description)
    
    plot_df <- formula_res@compareClusterResult %>% 
      # dplyr::group_by(brain_area, comparison, sig_reg, parent_term) %>%
            dplyr::group_by(brain_area, parent_term) %>%
      dplyr::filter(p.adjust == min(p.adjust)) %>%
      dplyr::filter(parent_sim_score == max(parent_sim_score)) %>%
      dplyr::ungroup() %>%
      dplyr::left_join(description_names, by = "parent_term") %>%
      dplyr::mutate(Description = mod_description) %>%
      # tidyr::separate(GeneRatio, into = c("gr1", "gr2"), convert = T) %>%
      # dplyr::mutate(geneRatio = gr1/gr2) %>%
      dplyr::group_by(Description) %>% 
      dplyr::mutate(min.padj = min(p.adjust)) %>% 
      dplyr::ungroup() %>%
      dplyr::arrange(desc(n),  brain_area, p.adjust) %>%
      dplyr::mutate(Description = factor(Description, levels = unique(Description)),
                    # sig_reg = factor(sig_reg, levels = c("Significantly Downregulated", "Significantly Upregulated")),
                    # test = factor(comparison, levels = c("PD vs Control")),
                    tissue = factor(brain_area, levels = c("ACG", "IPL", "MFG", "MTG")))
  }else{
    plot_df <- formula_res@compareClusterResult %>% 
      dplyr::rowwise() %>% 
      dplyr::mutate(Description = stringr::str_to_title(Description),
                    Description = paste0(strwrap(Description, width = 50, prefix = "\n", initial = ""), 
                                         collapse = "")) %>% 
      dplyr::group_by(Description) %>% 
      dplyr::mutate(n = n(),
                    min.padj = min(p.adjust)) %>% 
      dplyr::ungroup() %>%
      dplyr::arrange(desc(n), min.padj, brain_area, p.adjust) %>%
      dplyr::mutate(Description = factor(Description, levels = unique(Description)),
                    # sig_reg = factor(sig_reg, levels = c("Significantly Downregulated", "Significantly Upregulated")),
                    # test = factor(comparison, levels = c("PD vs Control")),
                    tissue = factor(brain_area, levels = c("ACG", "IPL", "MFG", "MTG")))
  }
}

drawPathwaysGO_tile <- function(formula_res_df, x='brain_area', x_expansion = c(0.5, 0.5), drop_facet = T, drop_x = F){
  plot<-formula_res_df %>%
    mutate(GeneRatio=DOSE::parse_ratio(GeneRatio)) %>%
   ggplot(aes(x = brain_area, y = Description, fill = -log10(p.adjust))) +
    geom_tile(color = "black", width = 1, show.legend = T) +
    geom_hline(yintercept = seq(0.5, length(plot_df$Description)), color="black", linewidth = 0.2) +
    geom_vline(xintercept = seq(0.5, length(plot_df[, x, T])+1), color="black", linewidth = 0.2) +
    scale_y_discrete(limits=rev, expand = expansion(add = x_expansion)) +
    scale_x_discrete(drop = drop_x, expand = expansion(add = x_expansion)) +
    scale_fill_viridis_c(name = "-log"[1][0]~"(FDR)", na.value = "white", option = "cividis") +
    coord_equal() +
    labs(x = "", y = "GO Term", fill = "hola") +
 # facet_wrap( ~ comparison)+
 #                  labeller = labeller(
 #                    sig_reg = c(
 #                      "Significantly Upregulated" = "Up-Reg.",
 #                      "Significantly Downregulated" = "Down-Reg."
 #                    )
 #                  ))+
    # custom_gg_theme +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          strip.text.x = element_text(size = 12, face = "plain"),
          text=element_text(size=14),
          panel.background = element_rect(fill = "white"),
          panel.grid.minor = element_blank(),
          panel.grid.major.y = element_blank(),
          panel.grid.major.x = element_blank(),
          panel.grid = element_blank())
    return(plot)
}

drawPathwaysGO_dot <- function(formula_res_df, x_expansion = c(0.75, 0.75), drop_facet = T, drop_x = F){  
    plot<-ggplot(formula_res_df %>% mutate(GeneRatio=DOSE::parse_ratio(GeneRatio)),
       aes(brain_area, Description, size=GeneRatio, colour=p.adjust))+
  geom_point()+
      # facet_wrap( ~ comparison)+
      #             labeller = labeller(
      #               sig_reg = c(
      #                 "Significantly Upregulated" = "Up-Reg.",
      #                 "Significantly Downregulated" = "Down-Reg."
      #               )
      #             ))+
  scale_y_discrete(limits=rev)+
  scale_size(range = c(2,8))+
  labs(x = "", y = "GO Term", fill = "hola") +
  # ggtitle(label = "enrichGO: Biological Process - Upregulated")+
  scale_color_gradientn(colours=c("#f7ca64", "#46bac2", "#7e62a3"),
                        trans = "log10", guide=guide_colorbar(reverse=TRUE, order=1))+
  theme(plot.title = element_text(size=18,vjust = 0.5),
        axis.text.x=element_text(angle=70, size=10, hjust = 0.95, vjust=0.95),
        axis.text.y = element_text(size=10),
        strip.text.x = element_text(size = 10),
        legend.key.size = unit(0.25, 'cm'), #change legend key size
        legend.key.height = unit(0.25, 'cm'), #change legend key height
        legend.key.width = unit(0.25, 'cm'), #change legend key widthsig_res
        legend.title = element_text(size=10), #change legend title font size
        legend.text = element_text(size=8))+
  DOSE::theme_dose(font.size = 10)+
  theme(axis.title.x = element_blank())
  
  return(plot)
}
```



## Table of significant clusters/GO IDs

Table of not-reduced and reduced pathways are shown below. They are ordered by the number of times the pathways are found across brain areas/comparisons and then adjusted p value.

### Not reduced

```{r go all paths no red dataframe, echo=F, eval=T, warning=F, message=FALSE,}

formula_res_merge_df <-generatePlotDataframe(formula_res_merge)

formula_res_merge_df %<>%
  # tidyr::separate(col = "Cluster", into = c("Tissue", "Regulation"), sep = "\\.", extra = "merge") %>% 
  dplyr::select(Description, brain_area, geneID, p.adjust, n, Ontology)

formula_res_merge_df %<>% arrange(p.adjust)

formula_res_merge_df %>% write_csv(path=file.path(results_path, "formula_res_merge_all.csv"))
```

### Reduced
```{r go all paths red dataframe, echo=F, eval=T, warning=F, message=FALSE}
formula_res_merge_red_df<-generatePlotDataframe(formula_res_merge_red)

formula_res_merge_red_df %<>% arrange(p.adjust) %>% write_csv(path=file.path(results_path, "formula_res_merge_reduced_all.csv"))

formula_res_merge_red_df %<>%
  # tidyr::separate(col = "Cluster", into = c("Tissue", "Regulation"), sep = "\\.", extra = "merge") %>%
  dplyr::select(Description, brain_area, geneID, p.adjust, n, Ontology)

formula_res_merge_red_df 

```

# GO Pathways

## No Reduction Performed
```{r, fig.width=12, fig.height=12}
formula_res_merge_df <- generatePlotDataframe(formula_res_merge)

plot_df <-
  formula_res_merge_df

if(nrow(plot_df) > 0){
  drawPathwaysGO_dot(plot_df)
}else{
  print("No pathways found in this scenario.")
}
```

## Reduction Performed

```{r, fig.width=15, fig.height=10}

formula_res_merge_red_df<- generatePlotDataframe(formula_res_merge_red)
plot_df <- formula_res_merge_red_df

if(nrow(plot_df) > 0){
  drawPathwaysGO_tile(plot_df)
}else{
  print("No pathways found in this scenario.")
}
```

# KEGG Pathways (No reduction) 

No pathways found

```{r, eval=F}
formula_res_KEGG_path <- file.path(results_path, "formula_KEGG.rds")

if(!file_test("-f", formula_res_KEGG_path)){
  formula_res_KEGG <- clusterProfiler::compareCluster(
    ENTREZID ~ brain_area,
    data = y, fun = enrichKEGG,
    universe = universe,
    organism = "hsa") 
  # No enrichment found
  # %>%
    # clusterProfiler::setReadable(OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
if(!is.null(formula_res_KEGG)){
  formula_res_KEGG@compareClusterResult %<>% dplyr::filter(Count >= count_threshold)

   
  # Save results
  formula_res_KEGG %>% saveRDS(formula_res_KEGG_path)
}else{
  formula_res_KEGG <- readRDS(formula_res_KEGG_path)
}

plot_df <- generatePlotDataframe(formula_res_KEGG)

if(nrow(plot_df) > 0){
  drawPathwaysGO_tile(plot_df)
}else{
  print("No pathways found in this scenario.")
}


}else{
  print("No pathways found in this scenario.")
}
```

# REACTOME enrichment analysis 

```{r,fig.width=10, fig.height=8}

formula_res_REACTOME_path <- file.path(results_path, "formula_REACTOME.rds")

if(!file_test("-f", formula_res_REACTOME_path)){
  
  formula_res_REACTOME <- clusterProfiler::compareCluster(
    ENTREZID ~ brain_area, 
    data = y, fun = "enrichPathway", 
    universe=universe,
    readable = T)
  
if(!is.null(formula_res_REACTOME)){
  formula_res_REACTOME@compareClusterResult %<>% dplyr::filter(Count >= count_threshold)
  
    # Save results
  formula_res_REACTOME %>% saveRDS(formula_res_REACTOME_path)
  
}else{
  formula_res_REACTOME <- readRDS(formula_res_REACTOME_path)
}

  plot_df <- generatePlotDataframe(formula_res_REACTOME)

if(nrow(plot_df) > 0){
  drawPathwaysGO_tile(plot_df)
}else{
  print("No pathways found in this scenario.")
}
  
  
  
}else{
  print("No pathways found in this scenario.")
}
  




```

# Conclusions



# Session Info

<button class="btn btn-primary" data-toggle="collapse" data-target="#BlockName"> Show/Hide </button>  
<div id="BlockName" class="collapse">  

```{r}
sessionInfo()
```

</div>
