---
title: |
  | PD Differential Gene Expression - PCA + Outliers
  | Hardy project dataset
subtitle: "Corrected identification script"
author: 
- name: "Guillermo Rocamora Pérez"
  affiliation: UCL - Version 1.5
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  bookdown::html_document2:
    figure_caption: yes
    code_folding: hide
    theme: paper
    highlight: haddock
    df_print: paged
    toc: true
    toc depth: 3
    toc_float: true
    number_sections: true
  md_document:
    variant: markdown_github
    toc: true
    number_sections: true
always_allow_html: true
link-citations: true
notes-after-punctuation: false
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/dependent/jhep-reports.csl
---

```{r setup, echo = T, warning=F, message=F}
## Load libraries
library(DESeq2)
library(patchwork)
library(foreach)
library(doParallel)
library(tidyverse)

## Define main path
main_path <- file.path("/home/grocamora/RytenLab-Research/18-DGE_limma_dream")# here::here()

## Load helper functions
source(file.path(main_path, "R/hf_graph_and_themes.R"))
source(file.path(main_path, "R/hf_reports.R"))
source(file.path(main_path, "R/run_stat_test_for_all_cols_get_res.R"))

## Set options
options(dplyr.summarise.inform = FALSE)
options(lifecycle_verbosity = "warning")
knitr::opts_chunk$set(echo = T, warning = F, message = F, 
                      out.width="100%", fig.align = "center", dpi = 100,
                      fig.width = 7.2, fig.height = 7.2)
```

```{r setup-theme, echo = F, results = "asis"}
## Custom sciRmdTheme with increased width.
## Please ask guillermorocamora@gmail.com for more details.
if(require(grpSciRmdTheme)){
  grpSciRmdTheme::set.theme(
    theme = "default",
    color = NULL,
    header.sticky = FALSE,
    list.group.icon = "arrow",
    font.family = "Arial",
    font.color = "black",
    header.color = "darkblue",
    figure.perc = 100
  )
}
## Custom styles for the output html
cat('
<style type="text/css">
.dataTables_scrollHeadInner{
  width:100% !important;
}
.dataTables_scrollHeadInner table{
  width:100% !important;
}
/*
.code-folding-btn {
  display: none;
}
*/
h3, .h3 {
  font-size: 22px!important;
}
h4, .h4 {
  font-size: 18px!important;
}
h5, .h5 {
  font-size: 16px!important;
}
body{
  font-size: 13px;
}

.tocify-subheader {
    text-indent: 15px;
    display: none;
    font-size: 12px;
}

.break5 .nav-tabs, .break4 .nav-tabs, .break3 .nav-tabs, .break2 .nav-tabs{
  display: grid;
  text-align: center;
  font-size: 14px;
  font-weight: bold;
}
.break5 .nav-tabs:before, .break4 .nav-tabs:before, .break3 .nav-tabs:before, .break2 .nav-tabs:before,
.break5 .nav-tabs:after, .break4 .nav-tabs:after, .break3 .nav-tabs:after, .break2 .nav-tabs:after {
  content: unset;
}

.break5 .nav-tabs{
  grid-template-columns: repeat(5, 1fr);
}
.break4 .nav-tabs{
  grid-template-columns: repeat(4, 1fr);
}
.break3 .nav-tabs{
  grid-template-columns: repeat(3, 1fr);
}
.break2 .nav-tabs{
  grid-template-columns: repeat(2, 1fr);
}

.breakB5 .nav-tabs{
  grid-template-columns: repeat(5, 1fr);
  font-size: 12px;
}
.breakB4 .nav-tabs{
  grid-template-columns: repeat(4, 1fr);
  font-size: 12px;
}
.breakB3 .nav-tabs{
  grid-template-columns: repeat(3, 1fr);
  font-size: 12px;
}
.breakB2 .nav-tabs{
  grid-template-columns: repeat(2, 1fr);
  font-size: 12px;
}

.stop-break .nav-tabs {
  display: flex;
  font-size: 13px;
  font-weight: normal;
}
.stop-break .nav-tabs:before,
.stop-break .nav-tabs:after {
  content: unset;
}
</style>')
```

# Background

Update of Jon's *PCA and PCA corrections report*. The covariates employed in this report are extracted from the script are extracted from:

> /home/grocamora/RytenLab-Research/18-DGE_limma_dream/Scripts/new_covariate_correction.R

The scripts updates the previous iteration by Jon with Aine's latest updates (as of 07/09/2023). The selected covariates are:

> GC_NC_40_59 + INTRONIC_BASES + PCT_UTR_BASES + Total_Sequences + gender + tss_up_1kb_tag_pct

As mentioned, this report relies on the previous execution of the *01-Hardy_covariate_identification.R* script, since it reads the output generated by it. Thus, we need to specify the path to input files, load the metadata and blacklisted genes:

```{r setup-variables, echo = T, results='hide'}
# Paths
main_path <- "/home/grocamora/RytenLab-Research/18-DGE_limma_dream"
results_path <- file.path(main_path, "results/Hardy_NewCovs")

txi.salmon_path <- file.path(main_path, "data/Hardy/txi.salmon.rds")
metadata_path <- file.path(results_path, "all_brain_areas_covariates.rds")
sample_metadata_path <- file.path(main_path, "metadata/Hardy/updated_metadata_w_PMI.csv")
sample_outliers_template_path <- file.path(results_path, "sample_outliers")
bxp_outliers_path <- file.path(results_path, "bxp_outliers.csv")

blacklist_genes <- file.path(main_path, "data", "blist_genes_no_vers.csv")
gene_map_path <- file.path(main_path, "data", "gencode_txid_to_geneid.txt")

# Load data
txi.salmon <- readRDS(txi.salmon_path)
cts <- txi.salmon$counts

metadata <- loadMetadataHardy(metadata_path, sample_metadata_path)

covariates <- c("GC_NC_40_59", "gender",
                "tss_up_1kb_tag_pct", "PCT_UTR_BASES",
                "Total_Sequences", "INTRONIC_BASES")

# Load blacklisted genes
blacklist_genes <- vroom::vroom(blacklist_genes, col_names = T, delim = ",")
blacklist_genes <- as.vector(unlist(blacklist_genes))

gencode_txid_to_geneid <- vroom::vroom(gene_map_path) %>%
  `colnames<-`(c("tx_id", "gene_id","gene_name", "description")) %>%
  dplyr::mutate(tx_id = sub("\\..+", "", tx_id),
                gene_id = sub("\\..+", "", gene_id))

# QC checks
all(colnames(cts) == metadata$bxp_id_full)
```

Additionally, we apply an expression threshold in which a $fpm>1$ in 50% of the samples is required for a gene to be employed in the analysis. We also remove the genes from the blacklisted regions.

```{r setup-dds, echo = T}
dds <- DESeq2::DESeqDataSetFromTximport(txi = txi.salmon,
                                        colData = metadata,
                                        design = reformulate(c(covariates, "Group")))

# Estimate library size correction scaling factors
dds <- DESeq2::estimateSizeFactors(dds)

# Remove blacklist genes
dds <- dds[!rownames(dds) %in% blacklist_genes, ]

# Identify genes that pass expression cutoff - then filter out genes that are
# extremely lowly expressed here, the standard cut-off set is fpm>1 in 50% or
# more of samples
isexpr <- rowSums(DESeq2::fpm(dds) > 1) >= 0.5 * ncol(dds)

# Compute log2 Fragments Per Million
quantlog.counts = log2(DESeq2::fpm(dds)[isexpr, ] + 1)
```

## Changelog {.tabset}

### v1.5 {-}

* Updated to R v4.3.0

### v1.4 {-}

* Fixed covariate identification mistake.

### v1.3 {-}

* Added additional results section with studies about brain region separation.

### v1.2 {-}

* Added two different selection of outliers: 18 outliers vs. 7 outliers.

* Added more graphs to assists with the decision of which outliers to remove.

### v1.1 {-}

* Added *WGCNA* approach to the outlier removal methods.

* Updated to use a total of 16 outliers.

### v1.0 {-}

* Initial report.

# PC Analysis by covariates

## Pre-correction {.tabset .break4}

First, we study if the covariates we selected group the samples in the PC plots. We select PCs until 85% of the variance is explained:

```{r pca-pre, echo = T, fig.height=5, fig.width=7.2}
pca <- prcomp(t(quantlog.counts))

highest_pc <- which(summary(pca)$importance[3, ] > 0.85)[1]

pca_out <- pca$x %>%
  .[, seq(highest_pc)] %>%
  tibble::as_tibble(rownames = "bxp_id_full")

var_explained = pca$sdev^2 %>% 
  tibble::tibble(PC = factor(1:length(.)), 
                 variance = .) %>% 
  dplyr::mutate(pct = variance/sum(variance)*100,
                pct_cum = cumsum(pct))

var_explained_plot = var_explained %>% 
  dplyr::filter(PC %in% c(1:(highest_pc*2))) %>% 
  ggplot(aes(x = PC)) +
  geom_col(aes(y = pct), color = "black", fill = ggsci::pal_jco()(3)[3]) +
  geom_line(aes(y = pct_cum, group = 1)) +
  geom_point(aes(y = pct_cum)) +
  geom_text(aes(label = round(pct, 2), y = pct + 3)) +
  scale_y_continuous(expand = expansion(), limits = c(0, 100), 
                     breaks = c(seq(0, 100, 25), 85)) +
  geom_hline(aes(yintercept = 85, linetype = "PC_limit"), linetype = "dashed",
             show.legend = F) +
  #scale_linetype_manual(name = "PC Threshold", values = "dashed", label = "85%") +
  labs(x = "Principal component", y = "% Variance explained") +
  custom_gg_theme + theme(legend.position = "right")

var_explained_plot
```

As shown in the figure, we only need 5 PCs to explain up to 85% of the variance.

```{r pca-pre-kw, echo = T}
pca_out %>%
  dplyr::left_join(metadata %>%
                     dplyr::select(bxp_id_full, Group),
                   by = "bxp_id_full") %>%
  kruskal.test(PC1 ~ Group, data = .)
```

```{r pca-pre-plot, results='asis', fig.width=9}
r <- sapply(c(covariates, "pd"), function(covariate){
  cat("### ", covariate, " {-}\n", sep = "")
  
  plot(plotPCgroup(pca_out, metadata, covariate))
  cat("\n\n")
})
```

## Post-correction {.tabset .break4}

We employ [`limma::removeBatchEffect`](https://rdrr.io/bioc/limma/man/removeBatchEffect.html) to covariate correct the salmon quantifications. In this example, our treatment matrix will consist of the *Group* variable.

```{r correction, echo = T, results='hide'}
covs_df <- metadata %>%
  dplyr::select(bxp_id_full, all_of(covariates), Group) %>%
  dplyr::mutate(across(where(is.numeric), ~ scale(.))) %>%
  tibble::column_to_rownames("bxp_id_full")

design <- model.matrix(reformulate(c(covariates, "Group")), data = covs_df)
design_treatment <- design[, grepl("(^Group)|(^\\(Intercept\\)$)", colnames(design))]
design_batch <- design[, !colnames(design) %in% colnames(design_treatment)]
design_batch_numeric <- design[, covs_df %>% dplyr::select(where(is.numeric)) %>% colnames()]
design_batch_categorical <- design[, !colnames(design) %in% c(colnames(design_treatment),
                                                              colnames(design_batch_numeric))]
# Test that rownames in design df == colnames in feature matrix
unique((quantlog.counts %>% colnames()) == (design %>% rownames())) 

# Apply removeBatchEffect to correct out the effect of the covariates
quantlog.counts_covar = limma::removeBatchEffect( # based on lmFit
  quantlog.counts,
  covariates = design_batch, # continuous/ numerical covariates here
  design = design_treatment)
```

After the covariate correction, we process with the same PC analysis as before.

```{r pca-post, echo = T, fig.height=5, fig.width=7.2}
pca_adj <- prcomp(t(quantlog.counts_covar))
# pca2 <- prcomp(t(quantlog.counts), center = TRUE, scale. = FALSE)

highest_pc_adj <- which(summary(pca_adj)$importance[3, ] > 0.85)[1]

pca_adj_out <- pca_adj$x %>%
  .[, seq(highest_pc_adj)] %>%
  tibble::as_tibble(rownames = "bxp_id_full")

var_explained_adj = pca_adj$sdev^2 %>% 
  tibble::tibble(PC = factor(1:length(.)), 
                 variance = .) %>% 
  dplyr::mutate(pct = variance/sum(variance)*100,
                pct_cum = cumsum(pct))

var_explained_adj_plot = var_explained_adj %>% 
  dplyr::filter(PC %in% c(1:(highest_pc_adj*1.1))) %>% 
  ggplot(aes(x = PC)) +
  geom_col(aes(y = pct), color = "black", fill = ggsci::pal_jco()(3)[3]) +
  geom_line(aes(y = pct_cum, group = 1)) +
  geom_point(aes(y = pct_cum)) +
  geom_text(aes(label = round(pct, 1), y = pct + 3), size = 3) +
  scale_y_continuous(expand = expansion(), limits = c(0, 100), 
                     breaks = c(seq(0, 100, 25), 85)) +
  geom_hline(aes(yintercept = 85, linetype = "PC_limit"), linetype = "dashed",
             show.legend = F) +
  #scale_linetype_manual(name = "PC Threshold", values = "dashed", label = "85%") +
  labs(x = "Principal component", y = "% Variance explained") +
  custom_gg_theme + theme(legend.position = "right")

var_explained_adj_plot
```

This time, we need 29 PCs to reach 85% of variance explained. Only the first 9 PCs will be shown in the following plots.

```{r pca-post-kw, echo = T}
pca_adj_out %>%
  dplyr::left_join(metadata %>%
                     dplyr::select(bxp_id_full, Group),
                   by = "bxp_id_full") %>%
  kruskal.test(PC1 ~ Group, data = .)
```

```{r pca-post-plot, results='asis', fig.height = 9, fig.width=12}
r <- sapply(c(covariates, "pd"), function(covariate){
  cat("### ", covariate, " {-}\n", sep = "")
  
  plot(plotPCgroup(pca_adj_out %>% dplyr::select(1:11), metadata, covariate))
  cat("\n\n")
})
```

## Brain region separation with control samples

An additional study is if the covariate correction helped to identify different brain regions in the control samples.

### All brain regions {.tabset .break2}

```{r brain-region-pca, echo = T}
pca_out_control <- pca_out %>%
  dplyr::left_join(metadata %>% dplyr::select(bxp_id_full, pd, brain_region), by = "bxp_id_full") %>%
  dplyr::filter(pd == "control") %>%
  dplyr::select(-pd, -brain_region)
pca_adj_out_control <- pca_adj_out %>%
  dplyr::left_join(metadata %>% dplyr::select(bxp_id_full, pd, brain_region), by = "bxp_id_full") %>%
  dplyr::filter(pd == "control") %>%
  dplyr::select(-pd, -brain_region)
```

Results show that no separation power is visually observed before and after the correction. 

#### Pre-correction {-}

```{r, brain-region-pre-plot, echo = F, fig.width=9}
plotPCgroup(pca_out_control, metadata, "brain_region")
```

#### Post-correction {-}

```{r brain-region-post-plot, echo = F, fig.height = 9, fig.width=12}
plotPCgroup(pca_adj_out_control %>% dplyr::select(1:19), metadata, "brain_region", by_pairs = F)
```

### IPL vs ACG {.tabset .break2}

```{r ipl-acg-pca, echo = T}
pca_out_control_filtered <- pca_out %>%
  dplyr::left_join(metadata %>% dplyr::select(bxp_id_full, pd, brain_region), by = "bxp_id_full") %>%
  dplyr::filter(pd == "control", brain_region %in% c("IPL", "ACG")) %>%
  dplyr::select(-pd, -brain_region)
pca_adj_out_control_filtered <- pca_adj_out %>%
  dplyr::left_join(metadata %>% dplyr::select(bxp_id_full, pd, brain_region), by = "bxp_id_full") %>%
  dplyr::filter(pd == "control", brain_region %in% c("IPL", "ACG")) %>%
  dplyr::select(-pd, -brain_region)
```

Results show that no separation power is visually observed before and after the correction. 

```{r ipl-acg-cluster-significance, echo = F, eval = F}
library(ClusterSignificance)
pca_brain_region <- pca_adj_out %>%
  dplyr::left_join(metadata %>% dplyr::select(bxp_id_full, pd, brain_region), by = "bxp_id_full") %>%
  dplyr::filter(pd == "control", brain_region %in% c("IPL", "ACG")) %>%
  dplyr::select(-pd, -bxp_id_full)

pcpMatrix <- as.matrix(pca_brain_region %>% dplyr::select(-brain_region))
rownames(pcpMatrix) <- pca_brain_region$brain_region

classes <- rownames(pcpMatrix)
prj <- pcp(pcpMatrix, classes)
plot(prj)
cl <- classify(prj)
plot(cl)

getData(cl, "AUC")

iterations <- 100
pe <- permute(mat = pcpMatrix, iter = iterations, classes = classes, projmethod = "pcp")
plot(pe)
pvalue(pe)
```

#### Pre-correction {-}

```{r ipl-acg-pre-plot, echo = F, fig.width=9}
plotPCgroup(pca_out_control_filtered, metadata, "brain_region")
```

#### Post-correction {-}

```{r ipl-acg-post-plot, echo = F, fig.height = 7.2, fig.width=9}
plotPCgroup(pca_adj_out_control_filtered %>% dplyr::select(1:9), metadata, "brain_region", by_pairs = F)
# plotPCgroup(pca_adj_out_control_filtered %>% dplyr::select(1:9), metadata, "brain_region", by_pairs = F) +
#   plot_annotation(
#   title = "PCs by brain region - Post corrected")
```

# Outlier removal

## PCA with Z-score {.tabset .break2}

```{r pca-outlier-function, echo = T}
getOutliersPC <- function(pca, var_explained = 85, max_pc = NULL, set_pc = NULL, z_score_threshold = 3){
  highest_pc <- which(summary(pca)$importance[3, ] > var_explained/100)[1]
  
  if(!is.null(set_pc)){
    max_outlier_pc <- set_pc
  }else if(!is.null(max_pc)){
    max_outlier_pc <- min(max_pc, highest_pc)
  }else{
    max_outlier_pc <- highest_pc
  }
  
  pc_outliers <- pca$x %>%
    .[, seq(max_outlier_pc)] %>%
    tibble::as_tibble(rownames = "bxp_id_full") %>%
    dplyr::left_join(metadata, by = "bxp_id_full") %>%
    dplyr::group_by(brain_region) %>%
    dplyr::mutate(across(paste0("PC", seq(max_outlier_pc)), list(zscore = ~ abs((. - mean(.))/sd(.))))) %>%
    dplyr::mutate(pc_number = max_outlier_pc,
                  z_score_pca_outlier = if_any(ends_with("zscore"), ~ . > z_score_threshold)) %>%
    dplyr::ungroup()
}
```

The PCA outlier detection approach extract the first *n* PCs and measures the $z-score$ of the samples for each PC. Samples are grouped by brain region. A sample is selected as an outlier if any of the their PCs' $z-score$ is greater than the threshold set at $>3$. Thus, the number of PCs employed to detect outliers is the main parameter to control for this approach.

For this study, we select PCs as a function of the total variance we want to explain. In the following plot, the X-axis represent the percentage of variance explained by the selected PCs, the size of the dots and the number inside them represent the number of PCs needed while the Y-axis represent the total number of outliers.

```{r pca-outlier-plot, echo = T, fig.height=4}
pc_outlier_summ <- parallel::mclapply(seq(35, 90, 5), mc.cores = (10), FUN = function(var_explained){
  pc_outliers <- getOutliersPC(pca_adj, var_explained)
  
  tibble::tibble(variance_explained = var_explained,
                 outliers_n = sum(pc_outliers$z_score_pca_outlier),
                 pc_number = unique(pc_outliers$pc_number))
}) %>% dplyr::bind_rows()

pc_outlier_summ %>%
  ggplot(aes(x = variance_explained, y = outliers_n)) + 
  geom_point(aes(size = pc_number)) +
  geom_line() +
  geom_vline(xintercept = c(65, 75), linetype = 2) +
  geom_text(aes(label = pc_number), vjust = 0.5, color = "white", fontface  = "bold", size = 3, show.legend = T) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)), breaks = seq(0, 50, 5), limits = c(0, NA)) +
  scale_x_continuous(expand = expansion(mult = c(0.1, 0.1)), breaks = seq(30, 90, 10)) +
  scale_size_continuous(range = c(4, 12), name = "Number of PCs") +
  labs(x = "% of variance explained", y = "Number of outliers") + 
  ggtitle("Number of outliers", subtitle = "By variance explained and number of PCs") + 
  custom_gg_theme +
  theme(legend.position = "right",
        legend.key.size = unit(0, 'cm'))

pc_outliers_6 <- getOutliersPC(pca_adj, set_pc = 6)
pc_outliers_10 <- getOutliersPC(pca_adj, set_pc = 10)
```

From the previous representation, it seems that two points stand out as relevant: i) 6 PCs with 4 outliers and ii) 10 PCs with 10 outliers. They precede changes in the trend. 

We can further study the $z-score$ distributions for these two different number of PCs:

```{r pca-outlier-plot-data, echo = F}
pc_outlier_plot_df <- dplyr::left_join(
  pc_outliers_6 %>%
    dplyr::rowwise() %>%
    dplyr::mutate(max_z6 = max(c_across(ends_with("zscore")))) %>%
    dplyr::select(bxp_id_full, max_z6) %>%
    dplyr::ungroup(),
  pc_outliers_10 %>%
    dplyr::rowwise() %>%
    dplyr::mutate(max_z10 = max(c_across(ends_with("zscore")))) %>%
    dplyr::select(bxp_id_full, max_z10) %>%
    dplyr::ungroup(),
  by = "bxp_id_full") %>%
  tidyr::pivot_longer(cols = -bxp_id_full, values_to = "zscore", names_to = "PCs")
```

### Plot for 6 PCs {-}

```{r, fig.height=4}
pc_outlier_plot <- pc_outlier_plot_df %>%
  dplyr::filter(PCs == "max_z6") %>%
  dplyr::mutate(outlier_type = ifelse(zscore > 3, "1", "0")) %>%
  ggplot(aes(x = bxp_id_full, y = zscore)) +
  geom_point(aes(color = outlier_type), show.legend = F, position = "jitter") +
  geom_hline(yintercept = c(3), linetype = 2) +
  scale_color_manual(values = c("black", "#6aa1fd")) +
  scale_y_continuous(limits = range(pc_outlier_plot_df$zscore)) +
  labs(x = "Samples", y = " Max. PCA Z-score") + 
  custom_gg_theme +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank())

ggExtra::ggMarginal(pc_outlier_plot, type = "histogram") 
```

### Plot for 10 PCs {-}

```{r, fig.height=4}
pc_outlier_plot <- pc_outlier_plot_df %>%
  dplyr::filter(PCs == "max_z10") %>%
  dplyr::mutate(outlier_type = ifelse(zscore > 3, "1", "0")) %>%
  ggplot(aes(x = bxp_id_full, y = zscore)) +
  geom_point(aes(color = outlier_type), show.legend = F, position = "jitter") +
  geom_hline(yintercept = c(3), linetype = 2) +
  scale_color_manual(values = c("black", "#6aa1fd")) +
  scale_y_continuous(limits = range(pc_outlier_plot_df$zscore)) +
  labs(x = "Samples", y = "Max. PCA Z-score") + 
  custom_gg_theme +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank())

ggExtra::ggMarginal(pc_outlier_plot, type = "histogram") 
```

## WGCNA sample connectivity and Z-score {.stop-break}

The WGCNA sample connectivity outlier approach also splits the samples by brain region and obtain the outliers from there. We need to decide on the $z-score$ threshold:

```{r}
getOutliersWGCNA <- function(quantlog, z_score_threshold = 2){
  mat <- quantlog
  
  # Calculate biweight midcorrelation matrix between samples
  bicor.mat <- WGCNA::bicor(mat)
  
  # Calculate signed adjacency matrix between samples
  normadj.mat <- (0.5 + 0.5*bicor.mat)^2
  
  # Calculate connectivity
  connectivity.mat <- WGCNA::fundamentalNetworkConcepts(normadj.mat)
  ku <- connectivity.mat$Connectivity
  z.ku <- (ku - mean(ku))/sqrt(var(ku))
  
  # Extract results
  sample_connectivity_res <- z.ku %>%
    stack() %>%
    dplyr::rename(bxp_id_full = ind, 
                  wgcna_connectivity_z_score = values) %>% 
    dplyr::mutate(z_score_wgcna_outlier = abs(wgcna_connectivity_z_score) > z_score_threshold) %>%
    dplyr::relocate(bxp_id_full, wgcna_connectivity_z_score)
  
  return(sample_connectivity_res)
}

getOutliersRegionWGCNA <- function(quantlog, metadata, z_score_threshold = 2){
  all_tissues <- metadata$brain_region %>% unique()
  
  sample_connectivity_res <- metadata$brain_region %>%
    unique() %>%
    parallel::mclapply(., mc.cores = length(.), FUN = function(tissue){
    bxp_ids <- metadata %>% dplyr::filter(brain_region == tissue) %>% dplyr::pull(bxp_id_full)
    
    mat <- quantlog[, bxp_ids]
    bicor.mat <- WGCNA::bicor(mat)
    
    # Calculate signed adjacency matrix between samples
    normadj.mat <- (0.5 + 0.5*bicor.mat)^2
    
    # Calculate connectivity
    connectivity.mat <- WGCNA::fundamentalNetworkConcepts(normadj.mat)
    ku <- connectivity.mat$Connectivity
    z.ku <- (ku - mean(ku))/sqrt(var(ku))
    
    # Extract results
    sample_connectivity_res <- z.ku %>%
      stack() %>%
      dplyr::rename(bxp_id_full = ind, 
                    wgcna_connectivity_z_score = values) %>% 
      dplyr::mutate(z_score_wgcna_outlier = abs(wgcna_connectivity_z_score) > z_score_threshold) %>%
      dplyr::relocate(bxp_id_full, wgcna_connectivity_z_score)
  }) %>% dplyr::bind_rows()

  return(sample_connectivity_res)
} 

wgcna_outliers_2 <- getOutliersRegionWGCNA(quantlog.counts_covar, metadata, z_score_threshold = 2)
wgcna_outliers_3 <- getOutliersRegionWGCNA(quantlog.counts_covar, metadata, z_score_threshold = 3)
```

```{r, fig.height=4}
wgcna_plot <- wgcna_outliers_3 %>%
  dplyr::mutate(outlier_type = case_when(abs(wgcna_connectivity_z_score) > 3 ~ "2",
                                         abs(wgcna_connectivity_z_score) > 2 ~ "1",
                                         T ~ "0")) %>%
  ggplot(aes(x = bxp_id_full, y = wgcna_connectivity_z_score)) + 
  geom_point(aes(color = outlier_type), show.legend = F) +
  scale_color_manual(values = c("black", "#66d487", "#6aa1fd")) +
  scale_y_continuous(breaks = seq(3, -6, -1)) +
  geom_hline(yintercept = c(-2, -3), linetype = 2) +
  labs(x = "Samples", y = "WGCNA Z-score") + 
  # ggtitle("Z-score of WGCNA connectivity approach") + 
  custom_gg_theme +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank())

ggExtra::ggMarginal(wgcna_plot, type = "histogram") 
```

We observe that with $z-score > 2$ we obtain 16 outliers, while with $z-score > 3$ we obtain 9 outliers.

## Combined approach {.tabset .break2}

If we combine the WGCNA outliers with both the 6 PCs and 10 PCs outliers, we observe the following Venn diagrams:

```{r, fig.width=8, fig.height=7}
wgcna_bxp_2 <- wgcna_outliers_2$bxp_id_full[wgcna_outliers_2$z_score_wgcna_outlier] %>% as.character()
wgcna_bxp_3 <- wgcna_outliers_3$bxp_id_full[wgcna_outliers_3$z_score_wgcna_outlier] %>% as.character()
pca_bxp_6 <- pc_outliers_6$bxp_id_full[pc_outliers_6$z_score_pca_outlier]
pca_bxp_10 <- pc_outliers_10$bxp_id_full[pc_outliers_10$z_score_pca_outlier]

outlier_list <- list("WGCNA_2" = wgcna_bxp_2, 
                     "WGCNA_3" = wgcna_bxp_3, 
                     "6 PCs" = pca_bxp_6, 
                     "10 PCs" = pca_bxp_10)

patchwork::wrap_plots(ggVennDiagram::ggVennDiagram(outlier_list[c("WGCNA_3", "6 PCs")]) + 
                        ggtitle("WGCNA with z-score > 3 & 8 PCs") +
                        theme(plot.margin = margin(0, 0, 0.5, 0, "cm")),
                      ggVennDiagram::ggVennDiagram(outlier_list[c("WGCNA_3", "10 PCs")]) + 
                        ggtitle("WGCNA with z-score > 3 & 10 PCs") +
                        theme(plot.margin = margin(0, 0, 0, 0, "cm")),
                      ggVennDiagram::ggVennDiagram(outlier_list[c("WGCNA_2", "6 PCs")]) + 
                        ggtitle("WGCNA with z-score > 2 & 8 PCs") +
                        theme(plot.margin = margin(0, 0, 0, 0, "cm")),
                      ggVennDiagram::ggVennDiagram(outlier_list[c("WGCNA_2", "10 PCs")]) + 
                        ggtitle("WGCNA with z-score > 2 & 10 PCs") +
                        theme(plot.margin = margin(0, 0, 0, 0, "cm")),
                      guides = "collect") & 
  scale_fill_gradient(limits=c(0, 16)) &
  theme(legend.position = "bottom")
```

Top left is the more lenient configuration while bottom right is the more stringent. A good middle approach good be to select bottom left, which includes 6 PCs and set the WGCNA z-score threshold to > 2.

An additional representation that may help decide on an outlier selection approach is to represent the $z-scores$ of all the samples that could be selected as outliers. The green bar represent the $z-score$ for the WGCNA approach, while the dots represent the $z-score$ for the different PCA methods. The PCA approach always require a $z-score>3$ to be considered a outlier, while the WGCNA may require $>2$ or $>3$:

```{r, fig.width=7.2, fig.height=5}
outlier_df <- tibble::tibble(bxp_id_full = outlier_list %>% unlist() %>% unique) %>%
  dplyr::left_join(wgcna_outliers_3 %>% 
                     dplyr::select(bxp_id_full, wgcna = wgcna_connectivity_z_score) %>%
                     dplyr::mutate(wgcna = abs(wgcna)),
                   by = "bxp_id_full") %>%
  dplyr::left_join(pc_outliers_6 %>% 
                     dplyr::rowwise() %>%
                     dplyr::mutate(pc_6_z_score = max(c_across(ends_with("zscore")))) %>%
                     dplyr::select(bxp_id_full, pc6 = pc_6_z_score),
                   by = "bxp_id_full") %>%
  dplyr::left_join(pc_outliers_10 %>% 
                     dplyr::rowwise() %>%
                     dplyr::mutate(pc_10_z_score = max(c_across(ends_with("zscore")))) %>%
                     dplyr::select(bxp_id_full, pc10 = pc_10_z_score),
                   by = "bxp_id_full") %>%
  dplyr::arrange(wgcna) %>%
  dplyr::mutate(bxp_id_full = factor(bxp_id_full, levels = unique(.$bxp_id_full))) %>%
  tidyr::pivot_longer(cols = -bxp_id_full, names_to = "method", values_to = "zscore") %>%
  dplyr::mutate(method = factor(method, levels = c("wgcna", "pc6", "pc10")))

outlier_df %>%
  {
    ggplot(., aes(x = bxp_id_full)) +
      # geom_tile(aes(y = 3, fill = ), stat = "identity", fill = "red", size = .5, width = 1, height = 7) +
      geom_col(aes(y = zscore, fill = method),
               data = . %>% dplyr::filter(method == "wgcna"),
               color = "black", width = 0.8, alpha = 1) +
      scale_fill_manual(values = "#83dc9e", label = "WGCNA", name = "Method") +
      ggnewscale::new_scale_fill() + 
      geom_point(aes(y = zscore, fill = method, shape = method, size = method), 
                 data = . %>% dplyr::filter(method != "wgcna")) +
      scale_size_manual(values = c(3, 2.15), label = c("pc6" = "6 PCs", "pc10" = "10 PCs"), name = "") +
      scale_shape_manual(values = c(21, 24), label = c("pc6" = "6 PCs", "pc10" = "10 PCs"), name = "") +
      scale_fill_manual(values = c("pc6" = "black", "pc10" = "#6aa1fd"), 
                        label = c("pc6" = "6 PCs", "pc10" = "10 PCs"), name = "") + 
      scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
      geom_hline(yintercept = c(2, 3), linetype = 2) +
      labs(x = "Samples", y = "Z-score") + 
      ggtitle("Z-Score for different methods") +
      coord_cartesian(ylim = c(0, 6)) +
      custom_gg_theme_clip +
      theme(axis.text.x = element_text(angle = 60, hjust = 1),
            legend.key = element_rect(fill = "white"))
  }
```

Based on these results, we decided to test two approaches:

### 16 outliers (6PCs all samples) {-}

All samples selected as outlier by any of the two methods when the $z-score$ threshold for *WGCNA* is $>2$ and up to 6 PCs are employed - all samples from the bottom left Venn diagram.

<details>
<summary>Expand to see selected samples.</summary>

```{r, echo = FALSE, fig.width=7.2, fig.height=5}
overlap_pc6 <- intersect(wgcna_bxp_2, pca_bxp_6)
outlier_df %>%
  dplyr::mutate(background = case_when(bxp_id_full %in% overlap_pc6 ~ "common",
                                       bxp_id_full %in% wgcna_bxp_2 ~ "wgcna",
                                       bxp_id_full %in% pca_bxp_6 ~ "pc6",
                                       T ~ "none")) %>%
  {
    ggplot(., aes(x = bxp_id_full)) +
      geom_tile(aes(y = 3, alpha = background, color = background, size = background, fill = background), 
                stat = "identity", 
                width = 1, height = 7, guides = guide_legend(nrow = 2),
                data = . %>% dplyr::distinct(bxp_id_full, background)) +
      scale_fill_manual(values = c("common" = "yellow", "wgcna" = "blue", 
                                                       "pc6" = "orange", "none" = "white")) +
      scale_alpha_manual(values = c("common" = 0.2, "wgcna" = 0.10, 
                                                        "pc6" = 0.2, "none" = 0)) +
      scale_color_manual(values = c("common" = "black", "wgcna" = "black", 
                                                        "pc6" = "black", "none" = "white")) +
      scale_size_manual(values = c("common" = 0.25, "wgcna" = 0.25, 
                                                       "pc6" = 0.25, "none" = 0)) +
      guides(fill = guide_legend(nrow = 2)) +
      ggnewscale::new_scale("size") +
      ggnewscale::new_scale("fill") +
      geom_col(aes(y = zscore, fill = method),
               data = . %>% dplyr::filter(method == "wgcna"),
               color = "black", width = 0.8, alpha = 1) +
      scale_fill_manual(values = "#83dc9e", label = "WGCNA", name = "Method") +
      ggnewscale::new_scale_fill() + 
      geom_point(aes(y = zscore, fill = method, shape = method, size = method), 
                 data = . %>% dplyr::filter(method != "wgcna")) +
      scale_size_manual(values = c(3, 2.15), label = c("pc6" = "6 PCs", "pc10" = "10 PCs"), name = "") +
      scale_shape_manual(values = c(21, 24), label = c("pc6" = "6 PCs", "pc10" = "10 PCs"), name = "") +
      scale_fill_manual(values = c("pc6" = "black", "pc10" = "#6aa1fd"), 
                        label = c("pc6" = "6 PCs", "pc10" = "10 PCs"), name = "") + 
      scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
      geom_hline(yintercept = c(2, 3), linetype = 2) +
      labs(x = "Samples", y = "Z-score") + 
      ggtitle("Z-Score for different methods") +
      coord_cartesian(ylim = c(0, 6)) +
      custom_gg_theme_clip +
      theme(axis.text.x = element_text(angle = 60, hjust = 1),
            legend.key = element_rect(fill = "white"))
  }
```

</details> \

The outliers selected are the following:

```{r}
bxp_outliers_6 <- unique(c(wgcna_bxp_2, pca_bxp_6))

sample_outliers_6 <- metadata %>%
  dplyr::left_join(wgcna_outliers_2 %>%
                     dplyr::select(bxp_id_full, 
                                   wgcna_zscore = wgcna_connectivity_z_score, 
                                   wgcna_outlier = z_score_wgcna_outlier), 
                   by = "bxp_id_full") %>%
  dplyr::left_join(pc_outliers_6 %>%
                     dplyr::rowwise() %>%
                     dplyr::mutate(max_z = max(c_across(ends_with("zscore")))) %>%
                     dplyr::select(bxp_id_full, 
                                   pca_zscore = max_z, 
                                   pca_outlier = z_score_pca_outlier), 
                   by = "bxp_id_full") %>%
  dplyr::mutate(outlier = ifelse(wgcna_outlier | pca_outlier, T, F))

sample_outliers_6 %>%
  dplyr::filter(bxp_id_full %in% bxp_outliers_6) %>%
  dplyr::select(bxp_id_full, case_id, brain_region, Group, 
                # wgcna_zscore, pca_zscore, 
                wgcna_outlier, pca_outlier) %>%
  dplyr::arrange(case_id, brain_region, Group) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(., 1))) %>%
  `colnames<-`(c("BXP ID", "Individual ID", "Brain region", "Group", 
                 # "WGCNA Zscore", "PCA Zscore", 
                 "WGCNA Outlier", "PCA outlier")) %>%
  printPrettyDf(style = "md")

number_outliers <- sum(sample_outliers_6$outlier == T)
sample_outliers_6 %>% saveRDS(paste0(sample_outliers_template_path, number_outliers, ".rds"))
```

After removing the outliers, the number of samples per group and brain region are:

```{r}
sample_outliers_6 %>% 
  dplyr::filter(!outlier) %>% 
  dplyr::select(brain_region, group) %>% 
  table() %>%
  `colnames<-`(c("Control", 
                 "Long without dementia", "Long with dementia", 
                 "Short without dementia", "Short with dementia")) %>%
  printPrettyDf(style = "md")
```

The PCA output with the outliers drawn:

```{r, fig.width=9, fig.height=6.5}
outlier_plots <- lapply(split(paste0("PC", seq(1, 8)), ceiling((1:8)/2)), function(a){
  plotZScorePCA(pc_outliers_10 %>% dplyr::mutate(z_score_outlier = bxp_id_full %in% bxp_outliers_6), a[1], a[2])
})

(outlier_plots[[1]] + outlier_plots[[2]])/
  (outlier_plots[[3]] + outlier_plots[[4]] ) +
  plot_layout(guides = "collect") &
  theme(legend.position = "top")
```

### 21 outliers (10 PCs intersection) {-}

All samples selected as outlier by any of the two methods when the $z-score$ threshold for *WGCNA* is $>2$ and up to 10 PCs are employed - all samples from the bottom left Venn diagram.

<details>
<summary>Expand see selected samples.</summary>

```{r, echo = FALSE, fig.width=7.2, fig.height=5}
overlap_pc10 <- intersect(wgcna_bxp_2, pca_bxp_10)
outlier_df %>%
  dplyr::mutate(background = case_when(bxp_id_full %in% overlap_pc10 ~ "common",
                                       bxp_id_full %in% wgcna_bxp_2 ~ "none",
                                       bxp_id_full %in% pca_bxp_10 ~ "none",
                                       T ~ "none")) %>%
  {
    ggplot(., aes(x = bxp_id_full)) +
      geom_tile(aes(y = 3, alpha = background, color = background, 
                    size = background, fill = background), 
                stat = "identity", 
                width = 1, height = 7, guides = guide_legend(nrow = 2),
                data = . %>% dplyr::distinct(bxp_id_full, background)) +
      scale_fill_manual(name = "Selection", values = c("common" = "yellow", "none" = "white"), labels = c("Common", "")) +
      scale_alpha_manual(name = "Selection", values = c("common" = 0.2, "none" = 0), labels = c("Common", "")) +
      scale_color_manual(name = "Selection", values = c("common" = "black", "none" = "white"), labels = c("Common", "")) +
      scale_size_manual(name = "Selection", values = c("common" = 0.25, "none" = 0), labels = c("Common", "")) +
      guides(fill = guide_legend(nrow = 1)) +
      ggnewscale::new_scale("size") +
      ggnewscale::new_scale("fill") +
      geom_col(aes(y = zscore, fill = method),
               data = . %>% dplyr::filter(method == "wgcna"),
               color = "black", width = 0.8, alpha = 1) +
      scale_fill_manual(values = "#83dc9e", label = "WGCNA", name = "Method") +
      ggnewscale::new_scale_fill() + 
      geom_point(aes(y = zscore, fill = method, shape = method, size = method), 
                 data = . %>% dplyr::filter(method != "wgcna")) +
      scale_size_manual(values = c(3, 2.15), label = c("pc6" = "6 PCs", "pc10" = "10 PCs"), name = "") +
      scale_shape_manual(values = c(21, 24), label = c("pc6" = "6 PCs", "pc10" = "10 PCs"), name = "") +
      scale_fill_manual(values = c("pc6" = "black", "pc10" = "#6aa1fd"), 
                        label = c("pc6" = "6 PCs", "pc10" = "10 PCs"), name = "") + 
      scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
      geom_hline(yintercept = c(2, 3), linetype = 2) +
      labs(x = "Samples", y = "Z-score") + 
      ggtitle("Z-Score for different methods") +
      coord_cartesian(ylim = c(0, 6)) +
      custom_gg_theme_clip +
      theme(axis.text.x = element_text(angle = 60, hjust = 1),
            legend.key = element_rect(fill = "white"))
  }
```

</details> \

The outliers selected are the following (in yellow the new ones with respect to 16 outliers):

```{r}
bxp_outliers_10 <- unique(c(wgcna_bxp_2, pca_bxp_10))

sample_outliers_10 <- metadata %>%
  dplyr::left_join(wgcna_outliers_2 %>%
                     dplyr::select(bxp_id_full, 
                                   wgcna_zscore = wgcna_connectivity_z_score, 
                                   wgcna_outlier = z_score_wgcna_outlier), 
                   by = "bxp_id_full") %>%
  dplyr::left_join(pc_outliers_10 %>%
                     dplyr::rowwise() %>%
                     dplyr::mutate(max_z = max(c_across(ends_with("zscore")))) %>%
                     dplyr::select(bxp_id_full, 
                                   pca_zscore = max_z, 
                                   pca_outlier = z_score_pca_outlier), 
                   by = "bxp_id_full") %>%
  dplyr::mutate(outlier = ifelse(wgcna_outlier | pca_outlier, T, F))

sample_outlier_table <- sample_outliers_10 %>%
  dplyr::filter(bxp_id_full %in% bxp_outliers_10) %>%
  dplyr::select(bxp_id_full, case_id, brain_region, Group, 
                # wgcna_zscore, pca_zscore, 
                wgcna_outlier, pca_outlier) %>%
  dplyr::arrange(case_id, brain_region, Group) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(., 1)))

sample_outlier_table  %>%
  `colnames<-`(c("BXP ID", "Individual ID", "Brain region", "Group", 
                 # "WGCNA Zscore", "PCA Zscore", 
                 "WGCNA Outlier", "PCA outlier")) %>%
  printPrettyDf(style = "md") %>% 
  kableExtra::row_spec(which(!(sample_outlier_table$bxp_id_full %in% bxp_outliers_6)), bold = T, 
                       color = "black", background = "#ffff99",)

number_outliers <- sum(sample_outliers_10$outlier == T)
sample_outliers_10 %>% saveRDS(paste0(sample_outliers_template_path, number_outliers, ".rds"))
```

```{r, eval = F}
a %>% dplyr::select(bxp_id_full, case_id, brain_region, Group, outlier_a = outlier) %>% 
  dplyr::left_join(b %>% dplyr::select(bxp_id_full, case_id, brain_region, Group, outlier_b = outlier)) %>% 
  dplyr::filter(outlier_a | outlier_b)
a_table <- a %>%
  dplyr::filter(outlier) %>% 
  dplyr::select(bxp_id_full, case_id, brain_region, Group, 
                wgcna_outlier, pca_outlier) %>%
  dplyr::arrange(case_id, brain_region, Group) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(., 1)))

b_table <- b %>%
  dplyr::filter(outlier) %>% 
  dplyr::select(bxp_id_full, case_id, brain_region, Group, 
                wgcna_outlier, pca_outlier) %>%
  dplyr::arrange(case_id, brain_region, Group) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(., 1)))
```


After removing the outliers, the number of samples per group and brain region are:

```{r}
sample_outliers_10 %>% 
  dplyr::filter(!outlier) %>%
  dplyr::select(brain_region, group) %>% 
  table() %>%
  `colnames<-`(c("Control", 
                 "Long without dementia", "Long with dementia", 
                 "Short without dementia", "Short with dementia")) %>%
  printPrettyDf(style = "md")
```

```{r, fig.width=12, fig.height=8}
outlier_plots <- lapply(split(paste0("PC", seq(1, 10)), ceiling((1:10)/2)), function(a){
  plotZScorePCA(pc_outliers_10 %>% dplyr::mutate(z_score_outlier = bxp_id_full %in% bxp_outliers_10), a[1], a[2])
})

patchwork::wrap_plots(outlier_plots, guides = "collect") &
  theme(legend.position = "top")
```

# PCs without 16 outliers {.tabset .break4}

```{r correction-clean, echo = T, results='hide'}
bxp_outliers <- unique(c(wgcna_bxp_2, pca_bxp_6))
data.frame(bxp_outliers) %>% readr::write_delim(bxp_outliers_path, delim = ",")

covs_clean_df <- metadata %>%
  dplyr::select(bxp_id_full, all_of(covariates), Group) %>%
  dplyr::mutate(across(where(is.numeric), ~ scale(.))) %>%
  dplyr::filter(!bxp_id_full %in% bxp_outliers) %>%
  tibble::column_to_rownames("bxp_id_full")

quantlog.counts_clean <- quantlog.counts[, !colnames(quantlog.counts) %in% bxp_outliers]

design <- model.matrix(reformulate(c(covariates, "Group")), data = covs_clean_df)
design_treatment <- design[, grepl("(^Group)|(^\\(Intercept\\)$)", colnames(design))]
design_batch <- design[, !colnames(design) %in% colnames(design_treatment)]
design_batch_numeric <- design[, covs_df %>% dplyr::select(where(is.numeric)) %>% colnames()]
design_batch_categorical <- design[, !colnames(design) %in% c(colnames(design_treatment),
                                                              colnames(design_batch_numeric))]
# Test that rownames in design df == colnames in feature matrix
unique((quantlog.counts_clean %>% colnames()) == (design %>% rownames())) 

# Apply removeBatchEffect to correct out the effect of the covariates
quantlog.counts_clean_covar = limma::removeBatchEffect( # based on lmFit
  quantlog.counts_clean,
  covariates = design_batch, # continuous/ numerical covariates here
  design = design_treatment)
```

In the end, we decided to go for the 16 outlier option (65% of variance explained). After the covariate correction and the outlier removal, the percentage of variance explained is the following:

```{r pca-post-clean, echo = T, fig.height=5, fig.width=7.2}
pca_adj <- prcomp(t(quantlog.counts_clean_covar))

highest_pc_adj <- which(summary(pca_adj)$importance[3, ] > 0.85)[1]

pca_adj_out <- pca_adj$x %>%
  .[, seq(40)] %>%
  tibble::as_tibble(rownames = "bxp_id_full")

var_explained_adj = pca_adj$sdev^2 %>% 
  tibble::tibble(PC = factor(1:length(.)), 
                 variance = .) %>% 
  dplyr::mutate(pct = variance/sum(variance)*100,
                pct_cum = cumsum(pct))

var_explained_adj_plot = var_explained_adj %>% 
  dplyr::filter(PC %in% c(1:(15*1.1))) %>% 
  ggplot(aes(x = PC)) +
  geom_col(aes(y = pct), color = "black", fill = ggsci::pal_jco()(3)[3]) +
  geom_line(aes(y = pct_cum, group = 1)) +
  geom_point(aes(y = pct_cum)) +
  geom_text(aes(label = round(pct, 1), y = pct + 3), size = 3) +
  scale_y_continuous(expand = expansion(), limits = c(0, 100), 
                     breaks = c(seq(0, 100, 25))) +
  # geom_hline(aes(yintercept = 85, linetype = "PC_limit"), linetype = "dashed", show.legend = F) +
  #scale_linetype_manual(name = "PC Threshold", values = "dashed", label = "85%") +
  labs(x = "Principal component", y = "% Variance explained") +
  custom_gg_theme + theme(legend.position = "right")

var_explained_adj_plot
```

```{r pca-post-clean-kw}
pca_adj_out %>%
  dplyr::left_join(metadata %>%
                     dplyr::select(bxp_id_full, Group),
                   by = "bxp_id_full") %>%
  kruskal.test(PC1 ~ Group, data = .)
```

```{r pca-post-clean-plot, results='asis', fig.height = 9, fig.width=12}
r <- sapply(c(covariates, "Group"), function(covariate){
  cat("## ", covariate, " {-}\n", sep = "")
  
  plot(plotPCgroup(pca_adj_out %>% dplyr::select(1:11), metadata, covariate))
  cat("\n\n")
})
```

# {.unlisted .unnumbered}

## PC Correlation and heatmap (Post correction) {.stop-break}

```{r, fig.width=12.5, fig.height=12, echo = T}
#### G: Highest PC to extract
highest_pc_name <- which(summary(pca_adj)$importance[3, ] > 0.85)[1]
highest_pc_name <- 20

#### A: spearman rho values
cors.r.numericVars = pca_adj %>%
  .[["x"]] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("bxp_id_full") %>%
  tibble::tibble() %>%
  .[, 1:(highest_pc_name+1)] %>%
  dplyr::left_join(x=.,
                   y=metadata %>% dplyr::select(bxp_id_full, where(is.numeric)),
                   by="bxp_id_full") %>%
  dplyr::select(-any_of(c("bxp_id_full", "sample_id"))) %>%
  dplyr::mutate(across(c(where(is.character)), as.factor),
                across(c(where(is.factor)), as.numeric)) %>%
  as.matrix() %>%
  Hmisc::rcorr(., type = "spearman") %>%
  .$r %>%
  as.data.frame() %>%
  dplyr::select(matches("PC\\d+")) %>%
  # J: need to get rid of some of the meta variables from picard that being with PCT
  dplyr::select(-contains("PCT")) %>%
  tibble::rownames_to_column("meta_var") %>%
  dplyr::filter(!grepl("PC\\d+", meta_var)) %>%
  tidyr::pivot_longer(2:ncol(.)) %>%
  dplyr::mutate(name = as.factor(as.numeric(gsub("PC", "", name)))) %>%
  dplyr::group_by(meta_var) %>%
  dplyr::mutate(max.PC.cor = max(value),
                mean.PC.cor = mean(value)) %>%
  dplyr::ungroup()

#### A: P-values
cors.p.numericVars = pca_adj %>%
  .[["x"]] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("bxp_id_full") %>%
  tibble::tibble() %>%
  .[, 1:(highest_pc_name+1)] %>%
  dplyr::left_join(x=.,
                   y=metadata %>% dplyr::select(bxp_id_full, where(is.numeric)),
                   by="bxp_id_full") %>%
  dplyr::select(-any_of(c("bxp_id_full", "sample_id"))) %>%
  dplyr::mutate(across(c(where(is.character)), as.factor),
                across(c(where(is.factor)), as.numeric)) %>%
  as.matrix() %>%
  Hmisc::rcorr(., type = "spearman") %>%
  .$P %>%
  as.data.frame() %>%
  dplyr::select(matches("PC\\d+")) %>%
  # J: need to get rid of some of the meta variables from picard that being with PCT
  dplyr::select(-contains("PCT")) %>%
  tibble::rownames_to_column("meta_var") %>%
  dplyr::filter(!grepl("PC\\d+", meta_var)) %>%
  tidyr::pivot_longer(2:ncol(.)) %>%
  dplyr::mutate(name = as.factor(as.numeric(gsub("PC", "", name))))

#### A: bind p and stat
cors.num = dplyr::left_join(
  x=cors.p.numericVars %>% dplyr::rename(p=value),
  y=cors.r.numericVars %>% dplyr::rename(stat=value),
  by=c("meta_var", "name")
) %>%
  dplyr::mutate(var_type = "continuous") %>%
  dplyr::group_by(meta_var) %>%
  dplyr::mutate(min.pvalue = min(p)) %>%
  dplyr::relocate(var_type, .after = last_col())

#### A: get PC/categorical correlations using K-W Chi-squared
#### G: added some regular expression to avoid issues with columns that are not
#### PCs.
cors.cat = pca_adj %>%
  .[["x"]] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("bxp_id_full") %>%
  tibble::tibble() %>%
  .[, 1:(highest_pc_name+1)] %>%
  dplyr::left_join(x=.,
                   y=metadata %>% dplyr::select(bxp_id_full, !where(is.numeric), -case_id),
                   by="bxp_id_full") %>%
  dplyr::select(-any_of(c("bxp_id_full", "sample_id", "group", "Group"))) %>%
  # A: run K-W test
  run_stat_test_for_all_cols_get_res(in_df=., stat_test="kruskal") %>%
  # A: filter so var1 contains the numerics, var2 contains the categorical
  dplyr::filter(grepl("PC\\d+", var1),
                !grepl("PC\\d+", var2)) %>%
  dplyr::select(-test_run) %>%
  dplyr::rename(name=var1, meta_var=var2) %>%
  dplyr::group_by(meta_var) %>%
  dplyr::mutate(max.PC.cor = max(stat),
                mean.PC.cor = mean(stat),
                min.pvalue = min(p)) %>%
  dplyr::mutate(name = as.factor(as.numeric(gsub("PC", "", name)))) %>%
  dplyr::ungroup() %>%
  dplyr::relocate(meta_var, name) %>%
  dplyr::mutate(var_type="categorical")

#### A: generate dataframe containing covariate, PC axis, rho & P-value
#### corresponding to PC-covariate, variance explained by each PC first, binding
#### cont (spearman) and cat (K-S) data to get df containing stat assessment for
#### all putative covariates
cors.all = dplyr::bind_rows(
  cors.num,
  cors.cat
) %>%
  dplyr::mutate(name = as.factor(name)) %>%
  dplyr::mutate(
    padj = p.adjust(p, method="fdr"),
    is.sig = case_when(
      padj<0.05 ~ stat,
      TRUE ~ NA_integer_
    )) %>%
  # A: add variance explained
  dplyr::left_join(
    x=.,
    y=summary(pca) %>%
      .$importance %>%
      as.data.frame() %>%
      t() %>%
      as.data.frame() %>%
      tibble::rownames_to_column("name") %>%
      janitor::clean_names() %>%
      dplyr::mutate(name = as.factor(gsub("PC", "", name))),
    by="name"
  ) %>%
  # A: calculate weighted correlation - weights PC-covariate correlation coefficient by the % variance explained by that PC
  dplyr::mutate(variance_weighted_correlation = proportion_of_variance*abs(stat)) %>%
  # A: add adjusted P-value - adjust for number of putative covariates
  dplyr::group_by(name) %>%
  dplyr::mutate(padj = p.adjust(p, "fdr")) %>%
  dplyr::ungroup()


#### A: generate PC-covariate heatmaps

#### A: plot PC-cov correlations
pca.cor.num = cors.all %>%
  dplyr::mutate(name = as.numeric(as.character(name))) %>%
  dplyr::filter(name %in% c(1:highest_pc_name)) %>%
  dplyr::filter(var_type=="continuous") %>%
  dplyr::arrange(name) %>%
  ggplot(data=., aes(x=name, y=reorder(meta_var, max.PC.cor), fill=stat)) +
  geom_tile() +
  geom_text(aes(label = round(is.sig, 2)), color = "black", size = 1.75) +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "#FFFFCC",
                       high = "#FF0000",
                       limits = c(-1, 1),
                       name = "Spearman's rho") +
  xlab("PC") +
  ylab("") +
  theme(axis.text.x = element_text(angle=0, vjust = 0.5, hjust=0.5),
        legend.position = "right") +
  scale_x_continuous(breaks=c(1:highest_pc_name)) +
  facet_wrap(~var_type, scales = "free")

pca.cor.cat = cors.all %>%
  dplyr::mutate(name = as.numeric(as.character(name))) %>%
  dplyr::filter(name %in% c(1:highest_pc_name)) %>%
  dplyr::filter(var_type=="categorical") %>%
  dplyr::arrange(name) %>%
  ggplot(data=., aes(x=name, y=reorder(meta_var, max.PC.cor), fill=stat)) +
  geom_tile() +
  geom_text(aes(label = ifelse(p<0.05, round(p, 2), "")), color = "black", size = 1.75) +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "#FFFFCC",
                       high = "#FF0000",
                       name = "K-W chi-squared") +
  xlab("PC") +
  ylab("") +
  theme(axis.text.x = element_text(angle=0, vjust = 0.5, hjust=0.5),
        legend.position = "right") +
  scale_x_continuous(breaks=c(1:highest_pc_name)) +
  facet_wrap(~var_type, scales = "free")

#### A: use patchwork to get cateorical and numeric/cont heatmaps in one figure
pca.cor.fig = (pca.cor.num / pca.cor.cat)
pca.cor.fig
```

# Additional results

## Brain separation {.tabset .break2}

Additionally, we wanted to study if there would be a brain area separation in the PC space after the covariate correction. Based on the collinearity plot shown before, we expect to see some separation power in the 16th PC.

```{r, results='asis', fig.height = 9, fig.width=12}
plotPCgroup(pca_adj_out %>% dplyr::select(1:19), metadata, "brain_region", by_pairs = F)
```

<!-- # Old code -->

<!-- ```{r, fig.height=4, eval = F} -->
<!-- # wgcna_bxp_2 <- wgcna_outliers_2$bxp_id_full[wgcna_outliers_2$z_score_wgcna_outlier] %>% as.character() -->
<!-- wgcna_bxp_3 <- wgcna_outliers_3$bxp_id_full[wgcna_outliers_3$z_score_wgcna_outlier] %>% as.character() -->

<!-- pca_bxp_8 <- pc_outliers_8$bxp_id_full[pc_outliers_8$z_score_pca_outlier] -->
<!-- pca_bxp_18 <- pc_outliers_8$bxp_id_full[pc_outliers_18$z_score_pca_outlier] -->

<!-- all_outliers <- wgcna_outliers_2 %>%  -->
<!--   dplyr::select(bxp_id_full, wgcna_z_2 = wgcna_connectivity_z_score, wgcna_outlier_2 = z_score_wgcna_outlier) -->

<!-- all_outliers <- dplyr::left_join( -->
<!--   wgcna_outliers_3 %>%  -->
<!--     dplyr::select(bxp_id_full, wgcna_z_3 = wgcna_connectivity_z_score, wgcna_outlier_3 = z_score_wgcna_outlier), -->
<!--   pc_outliers_8 %>% -->
<!--     dplyr::rowwise() %>% -->
<!--     dplyr::mutate(max_z = max(c_across(ends_with("zscore")))) %>% -->
<!--     dplyr::select(bxp_id_full, pca_z_8 = max_z, pca_outlier_8 = z_score_pca_outlier), -->
<!--   by = "bxp_id_full") %>% -->
<!--   dplyr::left_join(pc_outliers_18 %>% -->
<!--                      dplyr::rowwise() %>% -->
<!--                      dplyr::mutate(max_z = max(c_across(ends_with("zscore")))) %>% -->
<!--                      dplyr::select(bxp_id_full, pca_z_18 = max_z, pca_outlier_18 = z_score_pca_outlier), -->
<!--                    by = "bxp_id_full") %>% -->
<!--   dplyr::rowwise() %>% -->
<!--   dplyr::mutate(similarities = sum(wgcna_outlier_3, pca_outlier_8, pca_outlier_18)) -->


<!-- all_outliers %>% -->
<!--   dplyr::filter(wgcna_outlier_2 | wgcna_outlier_3 | pca_outlier_8 | pca_outlier_18) %>% -->
<!--   ggplot(aes(x = bxp_id_full)) +  -->
<!--   # geom_point(aes(y = abs(wgcna_z_2)), color = "black") +  -->
<!--   geom_point(aes(y = abs(wgcna_z_3)), color = "black") + -->
<!--   geom_point(aes(y = abs(pca_z_8)), color = "red") +  -->
<!--   geom_point(aes(y = abs(pca_z_18)), color = "blue") + -->
<!--   geom_hline(yintercept = 3) + -->
<!--   custom_gg_theme + -->
<!--   theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) -->
<!--   dplyr::filter(rowSums(. %>% dplyr::select(all_of("wgcna_z_2", "wgcna_z_3", "pca_z_8", "pca_z_18"))) >= 1) -->

<!-- all_outliers %>% -->
<!--   dplyr::filter(wgcna_outlier_2 | wgcna_outlier_3 | pca_outlier_8 | pca_outlier_18) %>% -->
<!--   ggplot(aes(x = reorder(bxp_id_full, similarities), y = similarities)) + -->
<!--   geom_col() -->

<!-- any_outliers <- unique(c(wgcna_bxp_2, wgcna_bxp_3, pca_bxp_8, pca_bxp_18)) -->

<!-- outlier_df <- tibble::tibble(bxp_id_full = any_outliers, -->
<!--                              wgcna_2 = any_outliers %in% wgcna_bxp_2, -->
<!--                              wgcna_3 = any_outliers %in% wgcna_bxp_3, -->
<!--                              pca_8 = any_outliers %in% pca_bxp_8, -->
<!--                              pca_18 = any_outliers %in% pca_bxp_18) -->

<!-- ggplot(wgcna_outliers_2, aes(x = wgcna_connectivity_z_score )) + -->
<!--   geom_histogram(bins = 30, fill = "steelblue", color = "black") + -->
<!--   # geom_density(fill = "steelblue", adjust = 1/2) + -->
<!--   geom_vline(xintercept = c(-2, -3), linetype = 2) +  -->
<!--   scale_y_continuous(expand = expansion(mult = c(0, 0.05))) + -->
<!--   custom_gg_theme -->
<!-- ``` -->


<!-- ```{r, eval = F} -->
<!-- r <- foreach(i = seq(60, 80, 5)) %do%{ -->
<!--   pc_outliers <- getOutliersPC(pca_adj, var_explained = i) %>% -->
<!--     dplyr::select(bxp_id_full, z_score_pca_outlier) %>% -->
<!--     dplyr::rename(!!(paste0("z_score_pca_outlier_", i)) := z_score_pca_outlier) -->

<!--   return(pc_outliers) -->
<!-- } %>% purrr::reduce(dplyr::left_join, by = "bxp_id_full") -->

<!-- wgcna_outliers <- getOutliersWGCNA(quantlog.counts_covar) -->

<!-- outliers <- r %>% -->
<!--   dplyr::left_join(wgcna_outliers %>% dplyr::select(bxp_id_full, z_score_wgcna_outlier),  -->
<!--                    by = "bxp_id_full") %>% -->
<!--   dplyr::relocate(z_score_wgcna_outlier) %>% -->
<!--   tibble::column_to_rownames("bxp_id_full") -->
<!--   # dplyr::mutate(across(everything(), ~ as.integer(.))) -->

<!-- bxp_ids <- rownames(outliers) -->
<!-- outliers_list <- apply(outliers, 2, function(x) bxp_ids[x]) -->

<!-- UpSetR::upset(outliers %>% dplyr::mutate(across(everything(), ~ as.integer(.))), keep.order = T) -->

<!-- UpSetR::upset(fromList(outliers_list), intersections = list( -->
<!--   list("z_score_wgcna_outlier", "z_score_pca_outlier_60"), -->
<!--   list("z_score_wgcna_outlier", "z_score_pca_outlier_65"), -->
<!--   list("z_score_wgcna_outlier", "z_score_pca_outlier_70"), -->
<!--   list("z_score_wgcna_outlier", "z_score_pca_outlier_75"), -->
<!--   list("z_score_wgcna_outlier", "z_score_pca_outlier_80")), -->
<!--   empty.intersections = "off") -->
<!-- ``` -->

<!-- Samples are grouped together by brain regions, and the z-scores of the first 10 PCs were calculated (until the variance explained by each PC is less than 2%). Samples with a $z−score>3$ of the selected PCs are considered outliers. -->

<!-- ```{r, fig.width=9, fig.height=6.5, eval = F} -->



<!--   pc_outliers <- pca_adj$x %>% -->
<!--     .[, seq(highest_pc_adj)] %>% -->
<!--     tibble::as_tibble(rownames = "bxp_id_full") %>% -->
<!--     dplyr::left_join(metadata, by = c("bxp_id_full")) %>% -->
<!--     dplyr::group_by(brain_region) %>% -->
<!--     # dplyr::mutate(across(matches("PC\\d+"), list(zscore = function(x) abs((x - mean(x))/sd(x))))) %>% -->
<!--     dplyr::mutate(across(paste0("PC", seq(1, max_outlier_pc)), ~ list(zscore = function(x) abs((x - mean(x))/sd(x))))) %>% -->
<!--     dplyr::mutate(z_score_outlier = if_any(ends_with("zscore"), ~ . > 3)) %>% -->
<!--     dplyr::ungroup() -->


<!--   #  -->
<!--   # %>% -->
<!--   #   patchwork::wrap_plots(guides = "collect", ncol = 3) & -->
<!--   #   theme(legend.position = "top") -->

<!--   pc_outliers %>% saveRDS(pc_outliers_path) -->
<!--   return(tibble::tibble(i = i, -->
<!--                         outliers_n = sum(pc_outliers$z_score_outlier), -->
<!--                         PCs = highest_pc_adj)) -->
<!-- } -->

<!-- r <- foreach(i = seq(35, 95, 5)) %do%{ -->
<!--   highest_pc_adj <- which(summary(pca_adj)$importance[3, ] > i/100)[1] -->
<!--   max_outlier_pc <- min(highest_pc_adj, highest_pc_adj) -->

<!--   pc_outliers <- pca_adj$x %>% -->
<!--     .[, seq(highest_pc_adj)] %>% -->
<!--     tibble::as_tibble(rownames = "bxp_id_full") %>% -->
<!--     dplyr::left_join(metadata, by = c("bxp_id_full")) %>% -->
<!--     dplyr::group_by(brain_region) %>% -->
<!--     # dplyr::mutate(across(matches("PC\\d+"), list(zscore = function(x) abs((x - mean(x))/sd(x))))) %>% -->
<!--     dplyr::mutate(across(paste0("PC", seq(1, max_outlier_pc)), list(zscore = function(x) abs((x - mean(x))/sd(x))))) %>% -->
<!--     dplyr::mutate(z_score_outlier = if_any(ends_with("zscore"), ~ . > 3)) %>% -->
<!--     dplyr::ungroup() -->

<!--   outlier_plots <- lapply(split(paste0("PC", seq(1, 10)), ceiling((1:10)/2)), function(a){ -->
<!--     plotZScorePCA(pc_outliers, a[1], a[2]) -->
<!--   })  -->

<!--   (outlier_plots[[1]] + outlier_plots[[2]])/ -->
<!--     (outlier_plots[[3]] + outlier_plots[[4]] + outlier_plots[[5]]) +  -->
<!--     plot_layout(guides = "collect") &  -->
<!--     theme(legend.position = "top") -->
<!--   #  -->
<!--   # %>% -->
<!--   #   patchwork::wrap_plots(guides = "collect", ncol = 3) & -->
<!--   #   theme(legend.position = "top") -->

<!--   pc_outliers %>% saveRDS(pc_outliers_path) -->
<!--   return(tibble::tibble(i = i, -->
<!--                         outliers_n = sum(pc_outliers$z_score_outlier), -->
<!--                         PCs = highest_pc_adj)) -->
<!-- } %>% dplyr::bind_rows() -->


<!-- ggplot(r, aes(x = i, y = outliers_n)) +  -->
<!--   geom_point() + -->
<!--   ggrepel::geom_text_repel(aes(label = PCs)) + -->
<!--   geom_line() + -->
<!--   labs(x = "% of variance explained", y = "Number of outliers") + -->
<!--   custom_gg_theme -->


<!-- ``` -->
